Copulas and their potential for ecology

### 2019-09-10 Paper.Rmd
new pedagog fig created, changes made to old pedagog fig

\begin{figure}[!h]
\begin{center}
\includegraphics[width=3.5in]{./Results/PedagogFig1.pdf}
\caption{Bivariate datasets showing diverse relationships between 
variables; some of these datasets nevertheless have the same Pearson (P)
or Spearman (S) correlation coefficients. Subscripts h and v 
in the axis labels stand for "horizontal" and "vertical".\label{fig_cop_pedag1}}
\end{center}
\end{figure}
<!--***DAN: When I cut panels out of this figure (saving them for use in a 
later figure), I also cut the caption, and here is the cut stuff for later use:
(C, D) Contours 
of probability density functions (pdf) of Clayton (C) and survival-Clayton 
(D) copulas with parameter 5 (see section on methods for Q1), together
with samples from the corresponding distributions. (E,F) Reflections
of C, D, respectively, about a vertical axis. (G) Contours of the
pdf of a normal copula (methods for Q1) with 
parameter chosen to
have the same Spearman correlation as A-D, together with a sample.
(H) The portion of the Spearman correlation within a band 
can be computed
for any band (methods for Q1) to describe how dependence
varies from one part of the two distributions to another. Diagonal lines
show two bands, the data in the left/lower band showing stronger
dependence than the right/upper band.
Subscripts h and v 
stand for "horizontal" and "vertical".
-->

<!--DONE ***DANr: reword this para so people don't think the ranks *are* the 
copulas, and to try to pre-defend against some of what ref 1 said. Also,
state that these are conceptual introductions only and are not formal,
a more comprehensive introduction is in xyz-->
Well-developed approaches do exist, however, and are 
applied widely in other fields, 
that provide a description of the dependence 
between two or more variables that can be mathematically proven 
to be complete [@nelsen2006_copula; @joe2014_dependence; @MaiScherer2017]. 
The approaches are based on 
the concept of the *copula*; copula approaches separate the information of a bivariate 
random variable into two distinct parts: the information in the 
marginal distributions (which says nothing about the dependence between the 
variables), and the remaining information (which is solely about
dependence). We here introduce some simple concepts, based on taking ranks,
that relate to copulas and give a conceptual 
flavor of copula ideas to be introduced 
more formally in the section on Background below. Given 
a sample $(x_i,y_i)$, $i=1,2,\ldots,n$ (e.g., 
Fig. \ref{fig_cop_pedag1}A-B), information about the structure of the dependence
between $x$ and $y$ can be separated from information contained 
in the marginal distributions by 
considering the bivariate plot of $u_i$ versus $v_i$, where
$u_i$ is the rank of $x_i$ in the set $x_j$ ($j=1,\ldots,n$), divided by
$n+1$; and $v_i$ is defined in the same way but using the $y_i$. The rank
of the smallest element in a set is $1$. The $u_i$ and 
$v_i$ are called *normalized ranks* of the $x_i$ and $y_i$; they relate to 
the empirical cummulative distribution function (cdf) of the $x_i$ and $y_i$,
respectively. See, for instance,
Fig. \ref{fig_cop_pedag1}C-D (the points), which show the normalized ranks of 
A-B, respectively. Ranking makes the marginal distributions of 
the component datasets uniform, isolating the 
dependence structure. Dependence structure and marginals can then be studied 
separately. Normalized rank plots such as Fig. \ref{fig_cop_pedag1}C-D
relate to copulas in a way to be described in the section on Background below, 
after a definition of copulas
is provided. Note that Pearson correlation, even though it is the most commonly
used measure of relationships between variables, is modified by monotonic
tranformation of the component variables (Appendix 
\ref{SM-pedagog1_details}, Fig. \ref{SM-fig_transf_pearson_spearman}),
and therefore reflects not only dependence information, but also
information on the marginals. Spearman and Kendall correlations, being rank-based, 
are not influenced by monotonic transformations of the component variables; 
they provide information solely on the dependence between those variables.

@@

<!--***DAN: I still feel like we have to make this more intuitive for people
at this point. Maybe a figure with two panels showing two copulas on
the left, and two panels shwoing two sets of marginals at the top,
and then four panels showing what you get when each of these combine?
Good idea. Steps: split the current Fig 1 so the new Fig 1 has the first four panels,
but C-D should only have the points. Then make a Fig 2 as above, and edit the
text in the above para to use that figure. 
The rest of the panels of the current Fig 1 will make a new Fig 3 later.-->

>>>>

\begin{figure}[!h]
\begin{center}
\includegraphics[width=3.5in]{./Results/PedagogFig1.pdf}
\caption{(A-D) Bivariate datasets showing diverse relationships between 
variables; some of these datasets nevertheless have the same Pearson (P)
or Spearman (S) correlation coefficients. C and D show normalized ranks of A, B, 
respectively. (E,F) Reflections
of C, D, respectively, about a vertical axis. Subscripts h and v 
in the axis labels stand for "horizontal" and "vertical".\label{fig_cop_pedag1}}
\end{center}
\end{figure}
<!--***DAN: When I cut panels out of this figure (saving them for use in a 
later figure), I also cut the caption, and here is the cut stuff for later use:
(C, D) Contours 
of probability density functions (pdf) of Clayton (C) and survival-Clayton 
(D) copulas with parameter 5 (see section on methods for Q1), together
with samples from the corresponding distributions. (E,F) Reflections
of C, D, respectively, about a vertical axis. (G) Contours of the
pdf of a normal copula (methods for Q1) with 
parameter chosen to
have the same Spearman correlation as A-D, together with a sample.
(H) The portion of the Spearman correlation within a band 
can be computed
for any band (methods for Q1) to describe how dependence
varies from one part of the two distributions to another. Diagonal lines
show two bands, the data in the left/lower band showing stronger
dependence than the right/upper band.
Subscripts h and v 
stand for "horizontal" and "vertical".
-->

<!--DONE ***DANr: reword this para so people don't think the ranks *are* the 
copulas, and to try to pre-defend against some of what ref 1 said. Also,
state that these are conceptual introductions only and are not formal,
a more comprehensive introduction is in xyz-->
Well-developed approaches do exist, however, and are 
applied widely in other fields, 
that provide a description of the dependence 
between two or more variables that can be mathematically proven 
to be complete [@nelsen2006_copula; @joe2014_dependence; @MaiScherer2017]. 
The approaches are based on 
the concept of the *copula*; copula approaches separate the information of a bivariate 
random variable into two distinct parts: the information in the 
marginal distributions (which says nothing about the dependence between the 
variables), and the remaining information (which is solely about
dependence). We here introduce some simple concepts, based on taking ranks,
that relate to copulas and give a conceptual 
flavor of copula ideas to be introduced 
more formally in the section on Background below. Given 
a sample $(x_i,y_i)$, $i=1,2,\ldots,n$ (e.g., 
Fig. \ref{fig_cop_pedag1}A-B), information about the structure of the dependence
between $x$ and $y$ can be separated from information contained 
in the marginal distributions by 
considering the bivariate plot of $u_i$ versus $v_i$, where
$u_i$ is the rank of $x_i$ in the set $x_j$ ($j=1,\ldots,n$), divided by
$n+1$; and $v_i$ is defined in the same way but using the $y_i$. The rank
of the smallest element in a set is $1$. The $u_i$ and 
$v_i$ are called *normalized ranks* of the $x_i$ and $y_i$; they relate to 
the empirical cummulative distribution function (cdf) of the $x_i$ and $y_i$,
respectively. See, for instance,
Fig. \ref{fig_cop_pedag1}C-D, which show the normalized ranks of 
A-B, respectively. Ranking makes the marginal distributions of 
the component datasets uniform, isolating the 
dependence structure. Dependence structure and marginals can then be studied 
separately. Normalized rank plots such as Fig. \ref{fig_cop_pedag1}C-D
relate to copulas in a way to be described in the section on Background below, 
after a definition of copulas
is provided. Note that Pearson correlation, even though it is the most commonly
used measure of relationships between variables, is modified by monotonic
tranformation of the component variables (Appendix 
\ref{SM-pedagog1_details}, Fig. \ref{SM-fig_transf_pearson_spearman}),
and therefore reflects not only dependence information, but also
information on the marginals. Spearman and Kendall correlations, being rank-based, 
are not influenced by monotonic transformations of the component variables; 
they provide information solely on the dependence between those variables.

@@

<!--***DAN: I still feel like we have to make this more intuitive for people
at this point. Maybe a figure with two panels showing two copulas on
the left, and two panels shwoing two sets of marginals at the top,
and then four panels showing what you get when each of these combine?
Good idea. Steps: 
DONE split the current Fig 1 so the new Fig 1 has the first 6 panels,
but C-D should only have the points. 
Then make a Fig 2 as above, and edit the text in the above para to use that figure. 
The rest of the panels of the current Fig 1 will make a new Fig 3 later.
-->
